\begin{enumerate}
  \item \label{trace:overrides-MyOuter-MyInner}Compute $\overrides((\text{``MyOuter''}, \text{``MyInner''}))$, the paths sharing the identity of $(\text{``MyOuter''}, \text{``MyInner''})$. It keeps $(\text{``MyOuter''}, \text{``MyInner''})$ and adds any same-label definition reached through the supers of its enclosing scope. Result: $\{(\text{``MyOuter''}, \text{``MyInner''})\}$.
  \item \label{trace:supers-MyOuter-MyInner}Compute $\supers((\text{``MyOuter''}, \text{``MyInner''}))$ using item~\ref{trace:overrides-MyOuter-MyInner}. It takes the reflexive-transitive closure of $\bases$ from $(\text{``MyOuter''}, \text{``MyInner''})$, pairing each reachable override with the inheritance site through which it is reached. Result: $\{((\text{``MyOuter''}), (\text{``MyOuter''}, \text{``MyInner''}))\}$.
  \item \label{trace:overrides-Has-outer}Compute $\overrides((\text{``HasMultipleOuters''}, \text{``outer''}))$, the paths sharing the identity of $(\text{``HasMultipleOuters''}, \text{``outer''})$ using item~\ref{trace:supers-MyOuter-MyInner}. It keeps $(\text{``HasMultipleOuters''}, \text{``outer''})$ and adds any same-label definition reached through the supers of its enclosing scope. Result: $\{(\text{``HasMultipleOuters''}, \text{``outer''}), (\text{``MyOuter''}, \text{``MyInner''}, \text{``outer''})\}$.
  \item \label{trace:bases-Has-outer}Compute $\bases((\text{``HasMultipleOuters''}, \text{``outer''}))$. $(\text{``HasMultipleOuters''}, \text{``outer''})$ carries no inheritance reference ($\inherits = \varnothing$), so it has no direct base. Result: $\varnothing$.
  \item \label{trace:bases-MyOuter-MyInner-outer}Compute $\bases((\text{``MyOuter''}, \text{``MyInner''}, \text{``outer''}))$. $(\text{``MyOuter''}, \text{``MyInner''}, \text{``outer''})$ carries the reference pair $(1, ())$ (take one step upward, then project the empty projection), whose target is obtained by item~\ref{trace:resolve-Has-MyOuter-MyInner-1}.
  \item \label{trace:this-Has-MyOuter-MyInner-1}Compute one $\this$ step: from the frontier $\{(\text{``HasMultipleOuters''})\}$, take the supers of each frontier path and keep those whose definition site is $(\text{``MyOuter''}, \text{``MyInner''})$, collecting their inheritance sites as the new frontier. Result: $\{(\text{``Object1''}), (\text{``Object2''})\}$.
  \item \label{trace:resolve-Has-MyOuter-MyInner-1}Compute $\resolve$ for the reference pair $(1, ())$ defined at $(\text{``MyOuter''}, \text{``MyInner''})$, reached from inheritance site $(\text{``HasMultipleOuters''})$ using item~\ref{trace:this-Has-MyOuter-MyInner-1}. It takes one step upward through $\this$ and then appends the projection. Result: $\{(\text{``Object1''}), (\text{``Object2''})\}$.
  \item \label{trace:resolve-0-MyOuter}Compute $\resolve$ for the reference pair $(0, (\text{``MyOuter''}))$ defined at $()$, reached from inheritance site $()$. It takes no step upward through $\this$ and then appends the projection. Result: $\{(\text{``MyOuter''})\}$.
  \item \label{trace:supers-Has-outer}Compute $\supers((\text{``HasMultipleOuters''}, \text{``outer''}))$ using items~\ref{trace:overrides-Has-outer}, \ref{trace:bases-Has-outer}, \ref{trace:bases-MyOuter-MyInner-outer}, \ref{trace:resolve-Has-MyOuter-MyInner-1} and~\ref{trace:resolve-0-MyOuter}. It takes the reflexive-transitive closure of $\bases$ from $(\text{``HasMultipleOuters''}, \text{``outer''})$, pairing each reachable override with the inheritance site through which it is reached. Result: $\{((), (\text{``MyOuter''})), ((), (\text{``Object1''})), ((), (\text{``Object2''})), ((\text{``HasMultipleOuters''}), (\text{``HasMultipleOuters''}, \text{``outer''})), ((\text{``HasMultipleOuters''}), (\text{``MyOuter''}, \text{``MyInner''}, \text{``outer''}))\}$.
  \item \label{trace:properties}Compute $\properties((\text{``HasMultipleOuters''}, \text{``outer''}))$ using item~\ref{trace:supers-Has-outer}. It unions the $\defines$ of every override path in that $\supers$ set. Only $(\text{``MyOuter''})$ defines a label, namely $\text{``MyInner''}$, reached through both $(\text{``Object1''})$ and $(\text{``Object2''})$; set union collapses the two routes. Result: $\{\text{``MyInner''}\}$.
\end{enumerate}
