% =====================================================================
%  The Fixed-Point over the Approximation Order
% ---------------------------------------------------------------------
%  Shared body. This file carries NO \documentclass: it is \input by two
%  entry points that supply the class (mimicking ../inheritance-calculus):
%    * preprint.tex   -- arXiv / preprint  (acmart, nonacm)
%    * submission.tex -- POPL submission   (acmart, review, anonymous)
%  Build both with: latexmk  (see latexmkrc)
% ---------------------------------------------------------------------
%  STATUS: working draft.
%  * Placeholders are marked \TODO{...} and render in red.
%  * Every bibliography entry is UNVERIFIED: author/title are best-effort
%    recollections; years, venues, volumes, and pages are marked [verify].
%    Do not cite from this file without checking each entry.
%  * Proofs are sketches; load-bearing gaps are flagged with \TODO.
% =====================================================================

%% acmart already loads amsmath, amssymb, amsthm, xcolor, and hyperref;
%% only mathtools needs to be added here.
\usepackage{mathtools}

% --- draft placeholder macro ------------------------------------------
\newcommand{\TODO}[1]{\textcolor{red}{[\textbf{TODO:} #1]}}
\newcommand{\citeTODO}[1]{\textcolor{red}{[\textbf{cite:} #1]}}

% --- general notation --------------------------------------------------
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Beh}{B}                         % behaviour functor B = M F
\newcommand{\rfp}{\varrho}                   % rational fixpoint
\newcommand{\behav}{\mathsf{beh}}            % behaviour map into the final coalgebra
\newcommand{\outm}{\mathsf{out}}             % structure map: the iteration of the equation morphism
\newcommand{\stepm}{e}                       % equation morphism (the map e)
\newcommand{\Delaym}{\mathsf{Delay}}
\newcommand{\Pow}{\mathcal{P}}
\newcommand{\iter}[1]{#1^{\dagger}}          % Elgot iteration

% --- lambda-instance notation (used only in the HNF bridge/examples) ---
\newcommand{\Sh}{\mathsf{Sh}}
\newcommand{\Lamsh}{\mathsf{Lam}}
\newcommand{\Appsh}{\mathsf{App}}
\newcommand{\Varsh}{\mathsf{Var}}
\newcommand{\dlam}{\mathord{\downarrow}\lambda}
\newcommand{\dhead}{\mathord{\downarrow}\mathsf{h}}
\newcommand{\darg}{\mathord{\downarrow}\mathsf{a}}
\newcommand{\subt}[2]{#1\!\downarrow\!#2}
\newcommand{\reduces}{\twoheadrightarrow}
\newcommand{\bred}{\rightarrow_{\beta}}
\newcommand{\present}{\mathtt{present}}

% --- theorem environments ---------------------------------------------
%% acmart provides theorem/lemma/proposition/corollary (acmplain) and
%% definition/example (acmdefinition), all sharing the `theorem' counter.
%% Only `remark' is missing, so we add it on the same shared counter.
\AtBeginDocument{%
  \theoremstyle{acmdefinition}%
  \newtheorem{remark}[theorem]{Remark}%
}

\begin{document}

\title{The Fixed-Point over the Approximation Order}

\author{Bo Yang}
\affiliation{%
  \institution{Figure AI Inc.}
  \city{San Jose}
  \state{California}
  \country{USA}
}
\email{yang-bo@yang-bo.com}
\thanks{This work was conducted independently prior to the author's employment at Figure AI.}

\begin{abstract}
Representing and transforming cyclic data in the pure $\lambda$-calculus normally requires an added recursion construct, such as a \texttt{letrec} or a $\mu$-binder, because reduction of a term that refers to itself does not terminate. Over the $\bot$-completed approximation order on a finitary signature, however, the behaviour functor is algebraically compact, so its initial algebra and final coalgebra coincide; the least and greatest fixpoints of a term's unfolding then agree, giving each term a unique fixpoint as its denotation, and what remains is to compute it. We present an algorithm that, for an arbitrary coalgebra over a finitary signature, computes this fixpoint by tabling. It halts precisely when the behaviour is rational, that is, when it has finitely many distinct subtrees. This is optimal: a non-rational behaviour has no finite representation, so no procedure that preserves the denotation halts on a larger class of inputs. Applied to the coalgebra that head reduction of the pure $\lambda$-calculus defines, the algorithm decides $\Omega$ as $\bot$ in finite time and represents $Y\,(\mathtt{cons}\;0)$, the cyclic stream of zeros, as a finite cyclic graph; an ordinary recursive function then transforms it, $\mathtt{map}\,\mathtt{succ}\,(Y\,(\mathtt{cons}\;0))$ returning the cyclic stream of ones as another finite cyclic graph, with no construct added to the calculus. An accompanying interpreter implements this case, evaluating any pure $\lambda$-term of rational behaviour to its unique fixpoint. We also give an encoding of pure Datalog into the $\lambda$-calculus; a Datalog program is function-free with a finite Herbrand base, so every encoded term has rational behaviour, and the interpreter computes the program's least Herbrand model.
\end{abstract}

\keywords{coalgebra, rational fixpoint, algebraic compactness, structure as identity,
  least fixpoint, rational trees, tabling, head reduction, Datalog,
  \texorpdfstring{$\lambda$-calculus}{lambda-calculus},
  \texorpdfstring{B\"ohm trees}{Bohm trees}}

\maketitle

% =====================================================================
\section{Introduction}\label{sec:intro}
% =====================================================================
A coalgebra is a state set $X$ with a structure map that, at each state, exposes one
layer of behaviour and names the successor states. Its \emph{behaviour} is the unfolding
of that map: the possibly infinite tree obtained by exposing layer after layer. We make one
observation about this unfolding and pursue its consequences. When the behaviour is
\emph{rational}, with finitely many distinct subtrees, it is the unfolding of a finite
cyclic graph, and that graph is computed by \emph{tabling}: identify a state with its own
behaviour, table the states seen so far, and when a state recurs while its behaviour is computed,
tie a back edge instead of unfolding forever.

Two things make this effective. The first is the form the input takes; the second is what makes its
cycles detectable at all.

First, the structure map need not be handed to us as a one-step observation. It may itself be
produced by an \emph{internal process}: from a state, perform some internal steps, possibly
none, until a layer of behaviour is exposed. We capture the internal process by an effect
monad $M$ and the exposed layer by a signature functor $F$, so the datum is a coalgebra
$\stepm\colon X \to M(F X + X)$ valued in the \emph{resumption} of $F$ over $M$
(Definition~\ref{def:itcoalg}): the $F X$ summand exposes a layer, the $X$ summand continues
silently, and $M$ carries the effect of the step. The one-step structure map
$\outm \colon X \to M(F X)$ is then recovered as the least solution of the iteration that runs
the silent steps (Definition~\ref{def:out}). Because it is a \emph{least} fixpoint, it ascends
from $\bot$, so a silent cycle, a state whose internal process returns to it with no layer
exposed, terminates as $\bot$ in finite time where the process alone diverges. Head
normalisation of the $\lambda$-calculus is exactly this situation, with weak-head reduction as
the internal process (Section~\ref{sec:bridge}).

Second, recognising that a state has recurred is a test of \emph{identity}. This is the word the
title turns on. The identity is decidable because the states are \emph{first-order}: they range
over graph nodes, never over a function space. The cycle fold is then nothing special; it is the
ordinary behaviour of tabling on state identity when the tabling core is the unfolding map. Where
the states are higher-order values, as in the na\"ive reading of the $\lambda$-calculus, the only
available identity is $\beta$-equality, which is undecidable, and neither memoisation nor cycle
folding is computable. Reading behaviour off a first-order coalgebra is what makes identity
decidable and the folding of cyclic structure effective.

\paragraph{Why this answers the cyclic-data question}
It is folklore that the pure $\lambda$-calculus cannot represent or manipulate cyclic data without
an added recursion construct (a \texttt{letrec}, a $\mu$-binder, observable sharing, mutable
references). The construction refutes this: the $\lambda$-calculus, read as an equation morphism,
is one instance among many, and on it a pure term such as $Y\,(\mathtt{cons}\;0)$ has a rational
behaviour, so tabling folds it into a finite circular graph and an ordinary recursive function over
it terminates, with nothing added to the calculus (Sections~\ref{sec:bridge}, \ref{sec:application}).
The capability is not special to the $\lambda$-calculus; it is a property of any equation morphism
with rational behaviour.

\paragraph{Contributions}
\begin{itemize}
  \item A reading of the behaviour of an equation morphism $\stepm \colon X \to M(F X + X)$ by
        tabling its one-step structure map over the reachable states, parametric in an effect monad
        $M$ (the internal process) and a signature functor $F$ (the exposed layer)
        (Sections~\ref{sec:domain}, \ref{sec:itcoalg}). The one-step map $\outm = \iter{\stepm}$ is
        the least solution of the silent iteration (Definition~\ref{def:out}), so divergence is
        decided as $\bot$ in finite time on the rational fragment.
  \item One fixpoint, and a graph for free (Sections~\ref{sec:itcoalg}, \ref{sec:readout}). The
        only fixpoint is the silent collapse $\outm = \iter{\stepm}$, an \emph{unguarded least}
        fixpoint resolving a silent cycle to $\bot$ (the bottom of $M$). Tabling $\outm$ over the
        reachable states yields the finite cyclic graph directly: its nodes are the states, and a
        cycle is automatic when $\outm$ names a state already tabled. No separate behaviour fixpoint is
        needed; the graph is the reachable sub-coalgebra $(R(M),\outm)$.
  \item Adequacy (Theorem~\ref{thm:adequacy}): the reachable sub-coalgebra $(R(M),\outm)$ unfolds to
        the behaviour of $\stepm$ up to silent steps, agreeing layer for layer with what the coalgebra
        enumerates. Termination (Theorem~\ref{thm:termination}): tabling halts iff finitely many
        states are reachable, and then the graph is a finite element of the rational fixpoint
        $\rfp\Beh$. Two independent finiteness conditions appear (Remark~\ref{rem:tworationalities}).
  \item Least equals greatest (Theorem~\ref{thm:unique}): because $\outm$ exposes a constructor or
        $\bot$, the behaviour functional is guarded, hence a contraction with a \emph{unique} fixpoint,
        so the least-fixpoint reading and the coinductive unfolding coincide and no least-versus-greatest
        choice arises. This is the operational face of algebraic compactness over the $\bot$-completed
        domain (Remark~\ref{rem:compact}); the least fixpoint is indispensable only at the silent
        collapse, where it decides divergence as $\bot$ (Remark~\ref{rem:leastessential}).
  \item A bridge (Section~\ref{sec:bridge}): head and weak-head normalisation of the pure
        $\lambda$-calculus are the instance $M = \Delaym$ (pure divergence) and $F$ the
        $\lambda$-shape signature. There $\outm$ is the weak-head shape relation, $\Omega$ reads out
        as $\bot$, and a guarded cyclic term reads out as a finite cyclic graph. Deterministic
        reduction strategies in general are the case $M = \Delaym$ (Remark~\ref{rem:determinism}).
  \item As an application, structural computation over cyclic data inside the pure
        $\lambda$-calculus (Section~\ref{sec:application}): an ordinary $\mathtt{map}$, with nothing
        cycle-aware in it, folds a circular list into a circular list, and a na\"ive walk terminates
        where head reduction diverges.
\end{itemize}

\paragraph{Supplementary material}
An executable implementation accompanies the paper: a first-order interpreter for the
$\lambda$-instance of Section~\ref{sec:bridge} ($\Varsh$/$\Lamsh$/$\Appsh$, no recursion binder), in
which a term's behaviour is read by tabling the weak-head structure map (Definition~\ref{def:shbridge}),
whose least-fixpoint silent collapse decides an unproductive cycle as $\bot$ and whose reachable
sub-coalgebra is the finite cyclic graph, printed with explicit back-references.\anon[ It is included
as supplementary material.]{ It is openly available
at \url{https://github.com/Atry/MIXINv2}~\cite{firstorderlambda}.} It realises the construction
directly: the silent collapse $\outm$ as a least fixpoint (with a finite-budget first iterate as a
policy flag), and the tabling that folds on structural identity, realised by interning so identity is
a pointer test. It also provides the head structure map of Remark~\ref{rem:boehm} as a second
least-fixpoint instance over the same carrier and behaviour map. The examples of
Sections~\ref{sec:overview}, \ref{sec:bridge}, and \ref{sec:application} are reproduced as tests:
$Y\,(\mathtt{cons}\;0)$ folds to a finite cyclic graph, $\Omega$ and $Y\,(\lambda x.x)$ read out as
$\bot$, the na\"ive walk terminates over the cyclic stream where head reduction loops, the
rationality dichotomy of Theorem~\ref{thm:termination} is exercised on a rational and a non-rational
stream, and the weak-head and head readings are checked to terminate together while differing on
$\lambda x.\,\Omega$ (Remark~\ref{rem:boehm}).

% =====================================================================
\section{Overview}\label{sec:overview}
% =====================================================================
Take the cyclic stream of zeros, $r = \mathtt{cons}\;0\;r$. As the behaviour of a coalgebra it is the
infinite tree $\mathtt{cons}\;0\;(\mathtt{cons}\;0\;\cdots)$, but it has only \emph{two} distinct
subtrees, $0$ and $r$ itself, so it is rational. Reading it off by tabling, the cell
$\mathtt{cons}\;0\;(-)$ is registered before its tail is followed; the tail returns to that very
cell, which is already in the table, so the tabling ties a back edge and stops. The result is a
finite graph, a node pointing to itself, that unfolds to the infinite stream. Any presentation of $r$
as an equation morphism with this behaviour gives the same finite graph, the $\lambda$-term
$Y\,(\mathtt{cons}\;0)$ among them (Section~\ref{sec:bridge}).

Now take a process that returns to a state \emph{without} ever exposing a layer: an
\emph{unproductive} cycle. There is nothing to tie a back edge to. This is the one place a fixpoint is
needed: the structure map $\outm$ is the least fixpoint of the silent iteration, so an unproductive
cycle, having nothing to fill its hole, settles at $\bot$. In the $\lambda$-instance this is reached
by a recursive binder ($\mathtt{letrec}\;x=x$) and equally by self-application,
$\Omega=(\lambda x.x\,x)(\lambda x.x\,x)$, which steps to itself with no constructor exposed, so
$\Omega = \bot$, decided in finite time where reduction loops forever.

So there is a single fixpoint, the silent collapse $\outm$. The guarded cycle above is not a second
fixpoint but a back edge the tabling ties for free: the tail already \emph{is} the registered cell, so
the cyclic graph is there in the coalgebra and computing the behaviour merely prints it. The guarded case folds,
the unproductive case bottoms out, and whether either terminates is governed by one property,
rationality, developed next.

% =====================================================================
\section{The behaviour domain}\label{sec:domain}
% =====================================================================
We fix the codomain of the behaviour map, the trees a coalgebra unfolds to, and the order along which the
fixpoints below ascend, before introducing the coalgebra that drives them.

\subsection{Signature and effect}
\begin{definition}[Signature functor]\label{def:signature}
A \emph{signature} is a finitary container functor $F\colon \Set \to \Set$,
$F X = \coprod_{\sigma \in \Sigma} X^{\mathit{ar}(\sigma)}$, for a set $\Sigma$ of operations with
finite arities $\mathit{ar}(\sigma) \in \mathbb{N}$. We require equality of \emph{shapes} (operations
together with their arity, i.e.\ the elements of $\Sigma$) to be decidable and hashable. The flat
domain of shapes is $F_{\bot} X = F X \cup \{\bot\}$, ordered by $\bot \sqsubseteq c$ for every $c$
and no other strict relation.
\end{definition}

\begin{definition}[Iteration-ready (Elgot) monad]\label{def:elgot}
An \emph{iteration-ready} monad is a monad $M$ on $\Set$ such that each $M A$ is a pointed
$\omega$-complete partial order with least element $\bot_{MA}$ (the \emph{divergent} computation), the
unit $\eta$ and Kleisli extension $(-)^{*}$ are $\omega$-continuous, and $\bot$ is preserved by
Kleisli composition (a divergent prefix diverges). Consequently every $f\colon X \to M(Y + X)$ has a
least \emph{iteration} $\iter{f}\colon X \to M Y$, the least fixpoint of
$g \mapsto \big(\,[\,\eta\circ\mathrm{inl},\,g\,]^{*}\big)\circ f$ in the pointwise order. This is the
order-theoretic presentation of a \emph{complete Elgot monad} \citeTODO{adamek-milius-velebil,
goncharov-elgot}.
\end{definition}

\begin{example}[Effects]\label{ex:effects}
$M = \Delaym$, the delay monad $\Delaym A = \nu Z.\,(A + Z)$ \citeTODO{capretta}, is the free
iteration-ready monad on the identity: $\bot$ is the infinite \texttt{later}, divergence. It models a
\emph{deterministic} process with no effect but possible non-termination, and is the case the
$\lambda$-calculus instance uses (Section~\ref{sec:bridge}). $M = \Pow$ (powerset) models
non-determinism, with $\bot = \varnothing$ and the iteration a set-valued least fixpoint; $M$ a
subdistribution monad models probabilistic choice; $M = S \Rightarrow (S \times -)$ models state.
Every definition and theorem below is stated for an arbitrary iteration-ready $M$.
\end{example}

\begin{definition}[Behaviour functor]\label{def:behfun}
The \emph{behaviour functor} is $\Beh = M\circ F$. A one-step coalgebra is a map $X \to \Beh X$. We
write $\nu\Beh$ for the final $\Beh$-coalgebra (the domain of behaviours) and $\rfp\Beh$ for its
\emph{rational fixpoint}, the colimit of the finite $\Beh$-coalgebras, equivalently the behaviours
with finitely many distinct subtrees \citeTODO{adamek-milius-velebil-rational}.
\end{definition}

\subsection{Trees, order, rationality}
For the order and the explicit tree presentation we spell out the deterministic skeleton; with a
non-trivial $M$ each node additionally carries the $M$-structure of its layer (a set of layers for
$M=\Pow$, a distribution for a probabilistic $M$, and so on), and the order and completeness lift
$M$-pointwise.

\begin{definition}[Trees]\label{def:domain}
A \emph{tree} is a partial $\Sigma$-labelling of the arity-respecting positions reachable by
descent, with a distinguished $\bot$ leaf for an exposed-nothing node. Formally a tree is a map
$t\colon \mathrm{dom}(t) \to \Sigma \cup \{\bot\}$ with $\mathrm{dom}(t)$ prefix-closed under the
descent into children and arity-respecting: a node labelled $\sigma$ has exactly $\mathit{ar}(\sigma)$
children and a node labelled $\bot$ has none. Let $D$ be the set of all finite or infinite such trees,
carrying the $M$-structure of Definition~\ref{def:behfun} at each node.
\end{definition}

\begin{definition}[Approximation order]\label{def:order}
For $t,u \in D$, $t \sqsubseteq u$ iff $\mathrm{dom}(t) \subseteq \mathrm{dom}(u)$ and for every
$p \in \mathrm{dom}(t)$ either $t(p) = u(p)$ or $t(p) = \bot$. Thus $u$ refines $t$ by replacing
$\bot$-leaves with subtrees (and, under a non-trivial $M$, by refining the $M$-structure
pointwise).
\end{definition}

\begin{proposition}[The codomain is a domain]\label{prop:cpo}
$(D,\sqsubseteq)$ is a pointed $\omega$-complete partial order with least element
$\bot_D$ (the single $\bot$ leaf), and least upper bounds of $\sqsubseteq$-directed families exist and
are computed nodewise. Proof in Appendix~\ref{app:domain}.
\end{proposition}

\begin{definition}[Rational behaviour]\label{def:rational}
$t \in D$ is \emph{rational} if its set of subtrees is finite; equivalently $t$ is the unfolding of a
finite $\Sigma$-labelled graph, equivalently $t \in \rfp\Beh$ (Definition~\ref{def:behfun}).
\end{definition}

% =====================================================================
\section{Equation morphisms and the structure map}\label{sec:itcoalg}
% =====================================================================
The development is parametric in the internal process that produces the structure map. We give it,
extract the one-step map by collapsing the silent steps, and record that the collapse is a least
fixpoint.

\begin{definition}[Equation morphism]\label{def:itcoalg}
An \emph{equation morphism} is a map
\[
  \stepm \colon X \longrightarrow M\big(F X + X\big),
\]
where $X$ is a set of \emph{states} with decidable, hashable equality. From a state, $\stepm$
performs one $M$-step whose result is either a layer of behaviour, $\mathrm{inl}(t)$ with $t \in F X$,
or a silent continuation, $\mathrm{inr}(x')$ with $x' \in X$. The $\mathrm{inl}$ alternative is
\emph{guarded} (a layer is exposed), the $\mathrm{inr}$ alternative \emph{unguarded} (the process
continues with no layer). Reachability is closed under both alternatives, so the children of an
exposed layer and every silent successor are again states.
\end{definition}

\begin{definition}[The structure map]\label{def:out}
The \emph{structure map} is the iteration that runs the silent steps to the next layer,
\[
  \outm \;:=\; \iter{\stepm} \colon X \longrightarrow M(F X),
\]
the least fixpoint (Definition~\ref{def:elgot}) of
$\Phi(g)(x) = \big(\,[\,\eta\circ\mathrm{inl},\,g\,]^{*}\big)(\stepm(x))$. Equivalently $\outm(x)$ runs
$\stepm$ from $x$, following $\mathrm{inr}$ continuations, and returns the first $\mathrm{inl}$ layer
under the accumulated $M$-effect; where the silent process never exposes a layer, $\outm(x)$ is the
divergent computation $\bot_{M(FX)}$.
\end{definition}

\begin{proposition}[The structure map is well-defined]\label{prop:outwd}
For every iteration-ready $M$ and every equation morphism $\stepm$, $\Phi$ is monotone and
$\omega$-continuous in the pointwise order on $X \to M(F X)$, so by Kleene
$\outm = \bigsqcup_{k} \Phi^{k}(\lambda x.\,\bot)$ exists and is the least fixpoint, and $\outm(x)$ is
the divergent computation exactly when the silent process from $x$ exposes no layer. Proof in
Appendix~\ref{app:out}.
\end{proposition}

\begin{remark}[Why the bottom is $\bot$, and why that is sound]\label{rem:bottom}
The iteration ascends from a bottom value, and that bottom is the divergent computation $\bot$ of $M$,
fresh and distinct from every exposed layer. This is what makes the resolution test exact: a state is
resolved precisely when its value is no longer $\bot$, so a silent cycle is recognised as soon as it
returns to a state already on the stack, and decided as $\bot$ in finite time. Reusing an exposed
value as the bottom would make a still-unresolved state indistinguishable from one that genuinely
computed that value, and a round could read a placeholder as a real value and reach a wrong fixpoint.
The only identity test, the one that lets the iteration recognise a cycle, is the decidable equality
of states, not an equality of the values they compute.
\end{remark}

The structure map turns the equation morphism $\stepm$ into an ordinary one-step coalgebra
$\outm \colon X \to \Beh X$ for $\Beh = M F$ (Definition~\ref{def:behfun}). The rest of the
development runs on $\outm$ and never re-inspects the silent steps. The next definition is the
positive presentation that makes $\Phi$ monotone above; like the rest, it is parametric in $M$ and
$F$.

\begin{remark}[Positivity]\label{rem:positivity}
The clauses of $\stepm$ are \emph{positive}: the choice between exposing a layer and continuing is a
positive disjunction over the $M$-result, never a negation-as-failure test such as ``no layer is
exposable''. Positivity is what makes $\Phi$ monotone, hence $\outm$ a least fixpoint by
Knaster--Tarski, so well-definedness of $\outm$ costs nothing beyond fixing the positive presentation.
The work of the paper is elsewhere (Sections~\ref{sec:readout}--\ref{sec:application}).
\end{remark}

% =====================================================================
\section{The rational graph: the reachable sub-coalgebra}\label{sec:readout}
% =====================================================================
The structure map is already an ordinary one-step coalgebra $\outm \colon X \to \Beh X$
(Definition~\ref{def:out}) whose successors are again states, so nothing further need be built: the
graph is the part of $(X,\outm)$ reachable from the root. Computing the behaviour is tabling $\outm$.

\begin{definition}[Reachable sub-coalgebra]\label{def:reach}
For a state $M$, the \emph{reachable set} $R(M) \subseteq X$ is the least set containing $M$ and
closed under the successors of $\outm$: if $x \in R(M)$ and $\outm(x)$ exposes a layer
$\sigma(x_1,\dots,x_k)$, then $x_1,\dots,x_k \in R(M)$ (for a branching $M$, every state in every
alternative of the layer). The \emph{graph} of $M$ is the sub-coalgebra $(R(M),\outm)$: nodes are the
reachable states, edges are $\outm$, and a node with $\outm(x)=\bot$ is the meaningless leaf.
\end{definition}

\begin{definition}[Reading the graph; the behaviour it denotes]\label{def:behreadout}
The graph $(R(M),\outm)$ is computed by \emph{tabling} $\outm$: keep a table of resolved states; from
$M$, compute $\outm(M)$ and, for each successor state, reuse its table entry or recurse and add one. A
successor already tabled closes a \emph{cycle}; a state with $\outm(x)=\bot$ is a leaf. The
\emph{behaviour} $\behav(M) \in \nu\Beh$ is what this graph denotes: the unfolding of $(R(M),\outm)$,
equivalently the unique $\Beh$-coalgebra morphism $X \to \nu\Beh$ at $M$ (finality). The behaviour is
never materialised; the graph is its finite presentation, and every question about it is answered on
the graph.
\end{definition}

\begin{remark}[One fixpoint; cycles for free]\label{rem:twofix}
There is only one fixpoint: the silent collapse $\outm = \iter{\stepm}$ (Definition~\ref{def:out}), an
unguarded least fixpoint that decides an unproductive cycle as $\bot$. Building the graph is not a
second fixpoint. A cycle appears the moment $\outm$ names a state already tabled, so the back edge is
structural, not tied by a separate computation, and ``reading out'' is just enumerating
$(R(M),\outm)$; printing it with explicit back-references is a traversal of that finite graph, again no
fixpoint. The whole algorithm is: table $\outm$.
\end{remark}

\begin{definition}[Guarded and unproductive states]\label{def:guarded}
A state $x$ is \emph{cyclic} if a state equal to it recurs strictly below it on some descent of
$\outm$. A cyclic $x$ is \emph{guarded} if $\outm(x)$ exposes a layer before the re-entry, and
\emph{unproductive} if $\outm(x) = \bot$, so the silent process of $x$ returns to it with no layer
exposed. A guarded cycle is a graph node pointing back to an earlier node, the finite representation of
a rational behaviour; an unproductive cycle is the $\bot$ leaf.
\end{definition}

% =====================================================================
\section{Adequacy and termination}\label{sec:adequacy}
% =====================================================================
\begin{theorem}[Adequacy]\label{thm:adequacy}
For every equation morphism $\stepm \colon X \to M(F X + X)$, the graph $(R(M),\outm)$
(Definition~\ref{def:reach}) unfolds to the behaviour of $\stepm$ up to silent steps: its unfolding
$\behav(M)$ is the unique $\Beh$-coalgebra morphism $X \to \nu\Beh$ through the structure map
$\outm = \iter{\stepm}$ (Definition~\ref{def:behreadout}), so the graph agrees layer for layer with
the unfolding the coalgebra enumerates once the silent steps are run. Proof in
Appendix~\ref{app:adequacy}.
\end{theorem}

\noindent
There is no normal-form translation to justify and no administrative bridge: once the silent steps are
collapsed into $\outm$, the graph is the behaviour's finite presentation essentially by construction,
and the only content is that the silent collapse $\iter{\stepm}$ presents the internal process
faithfully (Proposition~\ref{prop:outwd}).

\begin{theorem}[Termination]\label{thm:termination}
For every equation morphism with decidable state equality, tabling $\outm$ from $M$ halts if and
only if finitely many states are reachable from $M$ (the reachable set $R(M)$ is finite,
Definition~\ref{def:reach}), and in that case the graph $(R(M),\outm)$ is a finite cyclic graph whose
unfolding lies in the rational fixpoint $\rfp\Beh$ (Definition~\ref{def:rational}). When infinitely
many states are reachable, the tabling enumerates them and diverges. Proof in
Appendix~\ref{app:termination}.
\end{theorem}

\begin{remark}[Two finiteness conditions]\label{rem:tworationalities}
Two independent conditions govern termination, both with respect to structural state identity. The
silent collapse $\outm(x)$ (Definition~\ref{def:out}) halts iff the silent orbit from $x$ is finite,
returning to a state already seen rather than producing ever-new ones. The graph
(Definition~\ref{def:behreadout}) halts iff the reachable states $R(M)$ are finite. They are logically
independent: the silent orbit of a state may be infinite while the reachable states are finite, or
conversely. Either failure is genuine non-termination of the construction.
\end{remark}

\begin{remark}[Uniqueness does not characterise halting]\label{rem:notmunu}
One might hope the tabling halts exactly when the least and greatest fixpoint coincide. They always
coincide (Theorem~\ref{thm:unique} gives a unique fixpoint for \emph{every} term), yet the tabling does
not halt on every term, so $\mu=\nu$ cannot be the criterion. The witness is the stream of all
naturals, $\mathit{nats} = Y(\lambda s.\lambda n.\,\mathtt{cons}\,n\,(s\,(\mathtt{succ}\,n)))\,\underline{0}$:
it is guarded, so its behaviour is that unique fixpoint, a well-defined infinite stream, but the stream
is \emph{non-rational} (its tails are pairwise distinct), so it has no finite representation and the
tabling diverges. Halting is governed by rationality, which is strictly stronger than uniqueness: the
rational behaviours are a proper subclass of the (always unique) behaviours.
\end{remark}

\begin{theorem}[Finite-representation barrier; optimality]\label{thm:optimal}
By a \emph{finite representation} of a behaviour $t \in \nu\Beh$ we mean a finite $\Sigma$-graph whose
unfolding is $t$ (a first-order datum, to be distinguished from a finite \emph{generator}, a term that
reduces to $t$). The statement is for the deterministic skeleton; for a branching $M$ the labels carry
the $M$-structure and $\rfp\Beh$ is the corresponding rational fixpoint, and the argument lifts
unchanged. Then:
\begin{enumerate}
  \item \emph{(Barrier.)} $t$ has a finite representation iff $t$ is rational (Definition~\ref{def:rational},
        i.e.\ $t \in \rfp\Beh$); hence no procedure halts with a finite representation of a non-rational
        behaviour, even though every behaviour has a finite generator.
  \item \emph{(Soundness.)} Whenever the tabling halts it outputs a finite representation
        (Theorems~\ref{thm:termination}, \ref{thm:adequacy}), so it halts only on rational behaviours.
  \item \emph{(Attainment.)} The tabling halts iff finitely many positions are reached up to the identity
        it folds on. Under structural identity the halting set is a proper subclass of the rational
        behaviours, the gap being the dead-subterm witnesses ($Y\,F\,\underline{0}$ with
        $F=\lambda s.\lambda n.\,\mathtt{cons}\,0\,(s\,(\mathtt{succ}\,n))$); folding additionally any
        positions of equal behaviour enlarges it, and in the limit, folding all behaviourally equal
        positions, the tabling halts iff the behaviour is rational. By (1) no sound identity pushes the
        halting set beyond the rational fragment.
\end{enumerate}
\end{theorem}
\begin{proof}
\emph{(1)} A finite representation is a finite $\Sigma$-graph; its unfolding has at most one distinct
subtree per node, hence finitely many, so it is rational. Conversely a rational $t$, having finitely
many distinct subtrees, is the unfolding of the finite graph on those subtrees with an edge to each
immediate subtree of a root; this is the standard equivalence for regular trees \citeTODO{courcelle},
and the rational behaviours are exactly $\rfp\Beh$ (Definition~\ref{def:behfun}). Hence a non-rational
$t$ is the unfolding of no finite $\Sigma$-graph, so no procedure can output one. The generator clause
is immediate: the term $M$ is a finite generator of $\behav(M)$ for every $M$, rational or not. The
barrier is against any \emph{regular} (first-order, finitely presented) representation, each of which
unfolds to only rational trees; only a Turing-complete generator, such as the term itself, escapes it.

\emph{(2)} If the tabling halts then by Theorem~\ref{thm:termination} the reached positions are finite
and the result is a finite graph in $\rfp\Beh$, which by Theorem~\ref{thm:adequacy} unfolds to
$\behav(M)$; so a halt witnesses $\behav(M) \in \rfp\Beh$.

\emph{(3)} The tabling descends through positions up to the folding identity and tables them, so it
halts iff those classes are finite (Theorem~\ref{thm:termination}). Structural identity can keep
infinitely many classes while the behaviour is rational: in the cited witness every tail
$Y\,F\,\underline{k}$ is a distinct closed term, yet all denote the one constant-$0$ stream, which is
rational; so the halting set is a proper subclass of the rational behaviours. An identity that only
ever identifies positions of equal behaviour is sound (it changes no tree, Theorem~\ref{thm:adequacy})
and yields fewer classes, hence a larger halting set; for the coarsest such identity two positions are
identified exactly when their subtrees are equal, so the classes are the distinct subtrees of
$\behav(M)$, finite iff $\behav(M)$ is rational. That identity therefore halts iff rational, and by (1)
no sound identity can do better. That coarsest identity is effective on the rational fragment, where
each fold is decided by cycle-detecting bisimulation; the implementation here ships structural identity
instead, which sits strictly below it, the gap being the dead-subterm witnesses.
\end{proof}

\begin{remark}[The optimum is relative to the denotation]\label{rem:fixeddenotation}
The optimum in Theorem~\ref{thm:optimal} is among procedures that \emph{preserve the denotation}, the
tree. A fold changes no tree exactly when it identifies only behaviourally equal positions, so
behavioural equivalence is the coarsest denotation-preserving identity, and it is what attains the
rational ceiling (structural identity, fixed by the syntax, sits below it; behavioural equivalence,
relative to the chosen normalisation, sits at it). Folding more would identify positions of
\emph{different} behaviour: that is a coarser normalisation, a different tree, hence a different
denotation. One can then terminate on more, even on $\mathit{nats}$ by collapsing its distinct tails,
but one is computing a different semantics. Thus Theorem~\ref{thm:optimal}(1) is the barrier \emph{at a
fixed denotation}; termination beyond it is bought only by changing what is computed.
\end{remark}

\begin{remark}[Optimal semi-decision]\label{rem:semidec}
Rationality is undecidable: given a machine $T$, the term that emits the constant-$0$ stream while
simulating $T$ and switches to emitting $0,1,2,\dots$ once $T$ halts has a rational behaviour iff $T$
never halts, so rationality is co-r.e.-hard. Hence the terminating inputs cannot be pre-selected. The tabling is thus a \emph{sound semi-decision procedure}
for rationality: on the rational fragment (up to its identity, Theorem~\ref{thm:optimal}(3)) it halts
and returns the finite representation, and elsewhere it diverges, which by Theorem~\ref{thm:optimal}(1)
is the most any procedure can achieve.
\end{remark}

\begin{remark}[Determinism: reduction strategies]\label{rem:determinism}
When $M = \Delaym$, an equation morphism is a deterministic observe-or-step system
$X \to F X + X$ (one silent successor, divergence the only effect), the behaviour functor is $\Beh = F$
with $\nu\Beh$ the ordinary $F$-trees, and the silent collapse decides an unproductive cycle as the
single value $\bot$. This is the case of a \emph{deterministic reduction strategy}: head and
weak-head normalisation (Section~\ref{sec:bridge}), and equally call-by-name and call-by-value head
reduction, differ only in the choice of $F$ and of which redex the step contracts, never in the
$M = \Delaym$ shape. Richer $M$ (non-deterministic, probabilistic, stateful) leaves the architecture
intact but replaces the cycle-to-$\bot$ collapse with the corresponding $M$-least fixpoint, for
$M = \Pow$ a set-valued one.
\end{remark}

% =====================================================================
\section{The behaviour has a unique fixpoint: least equals greatest}\label{sec:unique}
% =====================================================================
The behaviour was introduced coinductively, as the unfolding of $\outm$ into $\nu\Beh$
(Definition~\ref{def:behreadout}). It is equally the order-theoretic least fixpoint of a behaviour
functional, and the two coincide: that functional is \emph{guarded}, so it has a \emph{unique}
fixpoint, and no least-versus-greatest choice arises. Once $\outm$ has exposed a constructor,
induction and coinduction agree.

\begin{definition}[Behaviour functional]\label{def:readoutfun}
The \emph{behaviour functional} $\Psi\colon (X \to D) \to (X \to D)$ reads one $\Beh$-layer off the
structure map and defers to its argument for the children:
\[
  \Psi(g)(n) =
  \begin{cases}
    \bot & \outm(n) = \bot,\\
    \sigma\big(g(n_1),\dots,g(n_k)\big) & \outm(n) \text{ exposes the layer } \sigma(n_1,\dots,n_k),
  \end{cases}
\]
the $M$-structure of the exposed layer (for a branching $M$) carried at the root, above the children.
By Proposition~\ref{prop:cpo} the pointwise order on $X \to D$ is a pointed $\omega$-CPO and $\Psi$ is
monotone and $\omega$-continuous, so $\operatorname{lfp}(\Psi) = \bigsqcup_{k} \Psi^{k}(\lambda n.\,\bot_D)$
exists.
\end{definition}

\begin{definition}[Tree metric]\label{def:metric}
For $t,u \in D$ put $\mathrm{d}(t,u) = 0$ if $t = u$, and otherwise $\mathrm{d}(t,u) = 2^{-\ell}$ where
$\ell$ is the least length $|p|$ of a position $p \in \mathrm{dom}(t) \cup \mathrm{dom}(u)$ at which $t$
and $u$ differ ($p$ lies in only one domain, or $t(p) \neq u(p)$). Lift it to $X \to D$ by
$\mathrm{d}(g,h) = \sup_{n} \mathrm{d}(g(n),h(n))$.
\end{definition}

\begin{proposition}[The tree space is complete]\label{prop:metric}
$(D,\mathrm{d})$ is a complete ultrametric space, and hence so is $(X \to D, \mathrm{d})$ under the
supremum metric \citeTODO{arnold-nivat}. A Cauchy sequence of trees stabilises at every fixed depth,
so it has a level-wise limit in $D$; the metric topology agrees with the $\omega$-CPO of
Proposition~\ref{prop:cpo} in that the $\mathrm{d}$-limit of a $\sqsubseteq$-increasing sequence is its
least upper bound.
\end{proposition}

\begin{lemma}[The behaviour functional is guarded]\label{lem:guarded}
$\Psi$ is a contraction with factor $\tfrac{1}{2}$:
$\mathrm{d}(\Psi(g),\Psi(h)) \le \tfrac{1}{2}\,\mathrm{d}(g,h)$.
\end{lemma}
\begin{proof}
Fix a state $n$. The root label of $\Psi(g)(n)$ is determined by $\outm(n)$ alone, independently of
$g$: it is $\bot$ when $\outm(n) = \bot$, and the operation $\sigma$ when $\outm(n)$ exposes
$\sigma(n_1,\dots,n_k)$. So $\Psi(g)(n)$ and $\Psi(h)(n)$ always share the root ($|p| = 0$) label and
can differ only strictly below it. If $\outm(n) = \bot$ or a nullary layer, they are equal and
$\mathrm{d}(\Psi(g)(n),\Psi(h)(n)) = 0$. Otherwise a shortest position at which they differ is one
descent step into some child $n_i$ followed by a shortest differing position of $g(n_i)$ and $h(n_i)$,
so its length is $1$ greater, giving
\[
  \mathrm{d}\big(\Psi(g)(n),\Psi(h)(n)\big) = \tfrac{1}{2}\,\max_{i}\,\mathrm{d}\big(g(n_i),h(n_i)\big)
  \le \tfrac{1}{2}\,\mathrm{d}(g,h).
\]
Taking the supremum over $n$ gives $\mathrm{d}(\Psi(g),\Psi(h)) \le \tfrac{1}{2}\,\mathrm{d}(g,h)$.
\end{proof}

\begin{theorem}[Least equals greatest]\label{thm:unique}
$\Psi$ has a unique fixpoint, and it is simultaneously
\begin{enumerate}
  \item the order-theoretic least fixpoint $\operatorname{lfp}(\Psi) = \bigsqcup_{k} \Psi^{k}(\lambda n.\,\bot_D)$, and
  \item the coinductive unfolding $\behav$, the unique $\Beh$-coalgebra morphism $X \to \nu\Beh$ over
        $\outm$ (Definition~\ref{def:behreadout}).
\end{enumerate}
Hence the least-fixpoint reading and the coinductive reading denote the same object: there is no
least-versus-greatest choice to make.
\end{theorem}
\begin{proof}
By Lemma~\ref{lem:guarded}, $\Psi$ is a contraction on the complete metric space $(X \to D, \mathrm{d})$
(Proposition~\ref{prop:metric}); by the Banach fixed-point theorem \citeTODO{banach} it has a unique
fixpoint $r$, and $\Psi^{k}(g_0) \to r$ from every starting point $g_0$.

\emph{(1)} Take $g_0 = \lambda n.\,\bot_D$. The iterates $\Psi^{k}(g_0)$ are $\sqsubseteq$-increasing
($\Psi$ is monotone and $g_0$ is least) and converge to $r$ in $\mathrm{d}$; the $\mathrm{d}$-limit of a
$\sqsubseteq$-increasing sequence is its least upper bound (Proposition~\ref{prop:metric}), so
$\bigsqcup_{k} \Psi^{k}(g_0) = r$, which by $\omega$-continuity is $\operatorname{lfp}(\Psi)$. Thus
$\operatorname{lfp}(\Psi) = r$.

\emph{(2)} A function $g$ is a fixpoint of $\Psi$ iff $g(n) = \sigma(g(n_1),\dots,g(n_k))$ whenever
$\outm(n)$ exposes $\sigma(n_1,\dots,n_k)$ and $g(n) = \bot$ when $\outm(n) = \bot$, which is exactly
the equation defining the $\Beh$-coalgebra morphism $X \to \nu\Beh$ over $\outm$. By finality that
morphism $\behav$ exists and is the unique such; being a fixpoint of $\Psi$, it equals $r$.

Therefore $\operatorname{lfp}(\Psi) = r = \behav$.
\end{proof}

\begin{remark}[Algebraic compactness]\label{rem:compact}
Theorem~\ref{thm:unique} is the operational face of a categorical fact. Over a \emph{discrete} carrier
the initial $\Beh$-algebra and the final $\Beh$-coalgebra differ, $\mu\Beh \neq \nu\Beh$ (for the
stream functor, the finite lists versus the finite-or-infinite ones), so there least and greatest
fixpoint genuinely diverge. The behaviour functional instead lives over the $\bot$-completed tree domain $D$, where a
locally continuous functor is \emph{algebraically compact}, $\mu\Beh \cong \nu\Beh$
\citeTODO{smyth-plotkin, freyd}, and the unique fixpoint of Theorem~\ref{thm:unique} is that one
common object. So ``least equals greatest'' is a property of the $\bot$-completion and the guarded
behaviour functional, \emph{not} of the discrete signature: completing with $\bot$ and reading guardedly is exactly
what collapses induction and coinduction.
\end{remark}

\begin{remark}[The parameter is the approximation order; the result is about computable ADTs]\label{rem:metalanguage}
What forces the unique fixpoint of Theorem~\ref{thm:unique} is the \emph{approximation order} on terms
(Definition~\ref{def:order}, equivalently its metric, Definition~\ref{def:metric}), under which the
guarded behaviour functional is a contraction; the iteration plays no role. This order is intrinsic to the
\emph{signature}, not to the $\lambda$-calculus: the Scott encoding supplies it automatically at the
implementation level (a constructor is observed, a divergence is $\bot$), but it can be defined
directly, with no reference to $\lambda$-terms, as the ideal completion of the finite partial
$\Sigma$-trees. Fixing it determines at once the order and the unique fixpoint of \emph{every} data
structure the calculus describes: a structure presented by a $\lambda$-term \emph{is} the tree read off
$\outm$, so it inherits both. Theorem~\ref{thm:unique} is therefore not really about the
$\lambda$-calculus. For any finitary signature $F$, every computable abstract data type, taken with its
approximation order, has a unique fixpoint, its initial algebra and final coalgebra coinciding
(Remark~\ref{rem:compact}); the $\lambda$-calculus is one realization of that order, not its source.
One never designates an iteration operator or selects a least solution \emph{for the data}; the least
fixpoint survives only one level down, in the evaluation, where the silent collapse gives a divergent
computation its forced value $\bot$ (Remark~\ref{rem:leastessential}).
\end{remark}

\begin{remark}[Where the least fixpoint is indispensable]\label{rem:leastessential}
Uniqueness is a property of the behaviour functional, after $\outm$ has exposed a constructor. The structure map
$\outm = \iter{\stepm}$ itself is \emph{unguarded}: a silent cycle returns with no constructor exposed,
so its defining operator (Definition~\ref{def:out}) is not contractive, and there least genuinely
differs from greatest. The least choice is the correct and the only computable one: $\Omega$ has
$\outm(\Omega) = \bot$ (least), whereas a greatest reading would assign it an arbitrary non-$\bot$
shape. So the construction is least-fixpoint throughout, and the single place the choice does real work
is the collapse of divergence to $\bot$; once a layer is exposed, the behaviour functional is guarded and least
equals greatest.
\end{remark}

% =====================================================================
\section{Bridge: head normalisation is an equation morphism}\label{sec:bridge}
% =====================================================================
Head normalisation of the pure $\lambda$-calculus is the instance $M = \Delaym$ (pure divergence,
Example~\ref{ex:effects}) with $F$ the $\lambda$-shape signature. It is the instance that bears out the
opening claim, so we give it in full; everything specialises the general development.

\begin{definition}[The $\lambda$-shape signature]\label{def:lamsig}
States are pure $\lambda$-terms in de Bruijn form (no $\alpha$-renaming), interned so that
structurally equal terms are one state. The signature is
\[
  F X = \big\{\Varsh_i\big\}_{i\in\mathbb{N}} \;+\; \big\{\Lamsh(x)\big\}_{x\in X} \;+\; \big\{\Appsh(x,x')\big\}_{x,x'\in X},
\]
a nullary $\Varsh_i$ per de Bruijn index, a unary $\Lamsh$, a binary $\Appsh$. There is \emph{no}
recursion binder; the repetition that tabling folds is produced by ordinary terms such as the
$Y$ combinator.
\end{definition}

\begin{definition}[Weak-head reduction as the equation morphism]\label{def:shbridge}
The equation morphism $\stepm \colon X \to \Delaym(F X + X)$ is weak-head reduction:
\[
  \stepm(M) =
  \begin{cases}
    \mathrm{now}\,\mathrm{inl}(\Varsh_i) & M = \Varsh_i\\
    \mathrm{now}\,\mathrm{inl}(\Lamsh(N)) & M = \Lamsh(\lambda x.N)\\
    \mathrm{now}\,\mathrm{inr}\big(\mathit{body}[x\mathbin{:=}a]\big) & M = \Appsh(f,a),\ f \text{ exposes } \Lamsh(\lambda x.\mathit{body})\\
    \mathrm{now}\,\mathrm{inl}(\Appsh(f,a)) & M = \Appsh(f,a),\ f \text{ exposes a rigid head.}
  \end{cases}
\]
A head redex is a silent step ($\mathrm{inr}$, contract and continue); a variable, an abstraction,
or an application with rigid head exposes a layer ($\mathrm{inl}$). The head $f$ is inspected through
its own structure map $\outm(f)$, so the clauses are mutually recursive through $\outm$, exactly the
silent iteration of Definition~\ref{def:out}.
\end{definition}

\begin{proposition}[The weak-head structure map is the weak-head shape]\label{prop:shisout}
For this $\stepm$, the structure map $\outm = \iter{\stepm}$ (Definition~\ref{def:out}) is the
weak-head shape relation $\Sh$: $\outm(M)$ is the outermost constructor of $M$ after weak head
reduction, and $\outm(M) = \bot$ exactly when weak head reduction of $M$ does not terminate. The
behaviour $\behav$ (Definition~\ref{def:behreadout}) is the L\'evy--Longo tree. Taking the structure
map to be head reduction \emph{under} $\lambda$ instead, a ``$\lambda$'' over a body with no head
normal form being itself $\bot$, gives the B\"ohm tree (Remark~\ref{rem:boehm}). Proof: specialise
Propositions~\ref{prop:outwd}, \ref{prop:cpo} and Theorem~\ref{thm:adequacy} to $M = \Delaym$.
\end{proposition}

\begin{remark}[Both readings, same convergence]\label{rem:boehm}
The weak-head and head structure maps are two instances over the one carrier, the one tabling, and
the one behaviour map, differing only in the $\lambda$ clause. They terminate on exactly the same terms:
head reduction performs eagerly, along the head spine, the very weak-head reductions the L\'evy--Longo
behaviour computation performs lazily down the $\lambda$-spine, so the reachable nodes and the reductions coincide.
They differ only denotationally, at a $\lambda$ whose body has no head normal form: L\'evy--Longo
keeps it (reading $\lambda x.\,\Omega$ as $\lambda x.\,\bot$), B\"ohm bottoms the whole term (reading
it as $\bot$). The choice of normalisation is therefore denotational, not a difference in
convergence. The accompanying implementation\anon[~(supplementary
material)]{~\cite{firstorderlambda}} realises both as least-fixpoint structure maps over the one
carrier and the one behaviour map, and checks both facts: $\lambda x.\,\Omega$ reads out as
$\lambda x.\,\bot$ under weak head and as $\bot$ under head, while both readings terminate on the same
terms and diverge on the same non-rational stream.
\end{remark}

\begin{example}[Guarded: the cyclic stream $Y\,(\mathtt{cons}\;0)$]\label{ex:cons}
Write $r = \mathtt{cons}\;0\;r$ as $Y\,(\mathtt{cons}\;0)$, with the Scott encoding
$\mathtt{cons} = \lambda h.\lambda t.\lambda c.\lambda n.\,c\,h\,t$. Then $\outm(r)$ exposes
$\lambda c.\lambda n.\,c\,0\,r'$ in finitely many silent steps without re-requiring $\outm(r)$: the
recursive occurrence enters only as data. The repeating cell is one interned state, so tabling
folds at the guarded re-entry,
\[
  \rho \;=\; \lambda c.\,\lambda n.\,(c\cdot 0)\cdot\rho,
\]
the finite representation of the rational L\'evy--Longo tree
$\lambda c.\lambda n.\,(c\,0)\,(\lambda c.\lambda n.\,(c\,0)\,(\cdots))$, a node pointing to itself,
where head reduction unfolds forever. No recursion binder is needed; $Y$ produces the repetition that
tabling folds.
\end{example}

\begin{example}[Unproductive: $\Omega$ and $Y\,(\lambda x.x)$]\label{ex:xx}
$\Omega = (\lambda x.x\,x)(\lambda x.x\,x)$ steps to itself with no constructor exposed, returning to
the same interned state; the silent collapse $\outm(\Omega)$ has nothing to expose, so $\outm(\Omega)
= \bot$ and $\Omega$ reads out as $\bot$, decided in finite time where reduction loops. The same holds
for $Y\,(\lambda x.x)$, the pure-$\lambda$ rendering of $\mathtt{letrec}\;x = x$.
\end{example}

% =====================================================================
\section{Applications}\label{sec:application}
% =====================================================================
\subsection{Decidability and dissolved problems}\label{sec:variants}
On the rational fragment the behaviour is a finite graph, so its equality is decidable, and a cluster
of problems that need added machinery in other settings does not arise.

\begin{corollary}[Decidable behaviour equality on the rational fragment]\label{cor:decidable}
Restricted to states from which finitely many states are reachable (Theorem~\ref{thm:termination}),
equality of behaviours $\behav(x) = \behav(y)$ is decidable. Each is presented by a finite cyclic
graph; the graphs are computable, since the silent collapse terminates at each reached state
(Proposition~\ref{prop:outwd}) and the reachable states close after finitely many steps, while state
identity is decidable (structural equality of states). Equality of two rational graphs is then decided
by the standard cycle-detecting bisimulation \citeTODO{courcelle}.
\end{corollary}

\begin{corollary}[Hard problems dissolved]\label{cor:dissolved}
Several problems that need added machinery elsewhere do not arise here. Observing sharing or cyclicity
in a pure language needs an observable-sharing primitive or a circular-program idiom
\citeTODO{gill, bird}; here sharing is state identity. Representing a cyclic structure needs an
explicit recursion binder or a cyclic $\lambda$-calculus \citeTODO{ariola-klop}; here the equation
morphism generates the repetition and tabling folds it, with no binder added
(Section~\ref{sec:bridge}). Deciding equality over cyclic structures needs a separate unification
discipline or recursive-type system \citeTODO{colmerauer, amadio-cardelli, brandt-henglein}; here it
is cycle detection over states (Corollary~\ref{cor:decidable}). Terminating a recursive query over a
cyclic datum needs a tabled evaluator \citeTODO{tamaki-sato, tabling-slg}; here the query's own
recurrences fall into the same states and fold (Section~\ref{sec:reach}).
\end{corollary}

\subsection{Structural computation over cyclic data}\label{sec:reach}
Constructing the cyclic datum is already the new step. The pure term $Y\,(\mathtt{cons}\;0)$ denotes
the infinite stream $\mathtt{cons}\;0\;(\mathtt{cons}\;0\;\cdots)$, with no finite form under head
reduction; up to identity it has finitely many states, so tabling $\outm$ gives it a finite
\emph{circular} representation, a cell whose tail points back to itself (Example~\ref{ex:cons}). A
circular linked list from a pure $\lambda$-term is what the standard semantics cannot build.

A recursive query over that datum is an ordinary $\lambda$-function, and the guarded/unproductive
split of Definition~\ref{def:guarded} governs the \emph{query} as it governs the data. The na\"ive
walk
\[
  \mathtt{reach} \;=\; Y\,\big(\lambda\mathit{self}.\,\lambda\mathit{lst}.\;
    \mathit{lst}\;(\lambda h.\lambda t.\,\mathit{self}\;t)\;\present\big)
\]
recurses down a Scott stream looking for $\mathtt{nil}$. Applied to $r = \mathtt{cons}\;0\;r$, each
recursive call builds $\mathtt{reach}\;t$; because $r$ has finitely many states, these fall into
finitely many states, so tabling folds the recursion. Head reduction, which remembers nothing,
builds a fresh application at every step and diverges; tabling folds at the repeated state and
returns a definite value (here $\bot$: no $\mathtt{nil}$ is reached, so the walk is itself an
unproductive cycle), in finite time. It is memoised for free, since identical states share their
subcomputations.

\paragraph{Mapping a cyclic list}
The guarded arm produces rather than decides. The textbook
$\mathtt{map}\,f = Y\,(\lambda\mathit{self}.\,\lambda\mathit{lst}.\;\mathit{lst}\;
(\lambda h.\lambda t.\,\mathtt{cons}\;(f\,h)\;(\mathit{self}\;t))\;\mathtt{nil})$ has nothing
cycle-aware in it. Its recursive call sits under a $\mathtt{cons}$, so it is guarded: applied to
$r = \mathtt{cons}\;0\;r$, the recursive application $\mathtt{map}\,f\;t$ re-enters the same state
$\mathtt{map}\,f\;r$, and tabling folds it. The result is a finite circular list
$\rho = \mathtt{cons}\;(f\,0)\;\rho$, every element transformed, where head reduction would map the
infinite stream forever. The fold is lazy knot-tying: each cell is registered before its tail is
recursed into, so the table supplies the back edge. On a finite list the same $\mathtt{map}$ is the
ordinary one. The implementation\anon[~(supplementary material)]{~\cite{firstorderlambda}} reproduces
both arms.

\subsection{Memoising arbitrary programs}\label{sec:memo}
More generally, $\behav$ is the denotation of a \emph{memoised} run of any program over the states.
Memoisation needs a comparable key on which to recognise a repeated call; first-order states supply
one, higher-order values do not (value equality of functions is undecidable). The tabling that computes the behaviour
is that memoisation: it is what lets a re-entry be recognised at all. Head reduction, which remembers
nothing, unfolds a cyclic state forever; tabling $\outm$ recognises the re-entry and folds. The two
are the unfolding and its limit, separated by how much they remember.

% =====================================================================
\section{Related work}\label{sec:related}
% =====================================================================
\paragraph{Coalgebra, final semantics, and rational fixpoints.}
Reading behaviour as the unique map into a final coalgebra is the coalgebraic method
\citeTODO{rutten, jacobs}; the finite cyclic graphs are the rational fixpoint between the initial
algebra and the final coalgebra \citeTODO{adamek-milius-velebil-rational}. Partition-refinement
minimisation of automata \citeTODO{hopcroft, paige-tarjan} folds on structural identity, but only for
automata, not for an arbitrary signature functor.

\paragraph{Iteration, the delay monad, and Elgot monads.}
Iteration in a (complete) Elgot monad
\citeTODO{adamek-milius-velebil, goncharov-elgot} solves guarded recursion abstractly, and for
$M = \Delaym$ the delay/partiality monad \citeTODO{capretta} makes divergence a value; but these
accounts fix the value categorically and neither compute it nor give it a finite representation on
the rational fragment. Generalised resumption monads $M(F(-) + -)$ \citeTODO{piróg-gibbons} supply the
type of such systems but no procedure over them.

\paragraph{Least Herbrand models and tabling.}
The least fixpoint of a monotone consequence operator is the least Herbrand
model of a definite specification \citeTODO{vanemden-kowalski}; plain resolution diverges on a cyclic
query, whereas a tabled evaluator \citeTODO{tamaki-sato, tabling-slg} computes the least fixpoint by
completing unproductive cycles instead of looping. Tabling there stays within logic programming and is
not carried to an arbitrary signature functor.

\paragraph{Rational trees and recursive types.}
Deciding structural queries over cyclic structures by cycle detection is long established: rational
(regular) trees and their algebra \citeTODO{courcelle} underlie cyclic unification
\citeTODO{colmerauer}, and equirecursive recursive types give terminating subtyping and equality by
detecting cycles in a regular tree \citeTODO{amadio-cardelli, brandt-henglein}. Each, however, handles
only the guarded-cycle case and is a separate discipline tied to one domain, unification or subtyping,
not a single uniform construction.

\paragraph{$\lambda$-trees and infinitary $\lambda$-calculus.}
The Berarducci, L\'evy--Longo, and B\"ohm trees form a hierarchy by their meaningless sets
\citeTODO{berarducci, levy-longo, barendregt}, situated uniformly by the infinitary
$\lambda$-calculus \citeTODO{kennaway-klop-sleep-devries}; they are studied as infinite normal forms,
with no finite representation of the rational ones. The Berarducci tree retains $\bot$-headed
applications and is non-monotone, so it admits no least-fixpoint presentation. The lazy
$\lambda$-calculus \citeTODO{abramsky-ong} observes weak head normal forms but does not fold the trees
it produces.

\paragraph{Cyclic $\lambda$-terms, sharing, and recursion.}
Graph reduction \citeTODO{wadsworth, turner-combinators} evaluates $\lambda$-terms over (possibly
cyclic) graphs, and term graph rewriting \citeTODO{barendregt-term-graph, kennaway-graph-adequacy}
relates finite cyclic graphs to the rational trees they unfold to; but a cycle is tied only where
recursion is primitive, a \texttt{letrec} or a built-in fixpoint, whereas a pure fixpoint combinator
has its body copied at each $\beta$-step and unfolds into an unbounded spine. Cyclic graphs / explicit
recursion \citeTODO{ariola-klop} and call-by-need \citeTODO{ariola-felleisen} likewise provide sharing
only as an added construct, and a pure functional language cannot even observe sharing or cyclicity
without an addition such as observable sharing \citeTODO{gill} or a circular-program idiom
\citeTODO{bird}. Across all of these, an ordinary recursive function over a cyclic list, such as a
$\mathtt{map}$, reduces to an unbounded spine rather than to a cycle.

\paragraph{Higher-order model checking and defunctionalization.}
Deciding properties of trees generated by higher-order programs \citeTODO{ong-hors, kobayashi}
answers queries about $\lambda$-generated infinite trees, but proceeds by automata and games over
B\"ohm-tree-like denotations and produces no finite representation of the tree itself.
Defunctionalization \citeTODO{reynolds} compiles a higher-order program into a first-order
interpreter, a syntactic translation that yields no semantic \emph{object}.

\paragraph{A common limitation.}
Terminating structural computation over cyclic data by cycle detection is decades old in logic
programming \citeTODO{colmerauer, tamaki-sato} and recursive-type theory \citeTODO{amadio-cardelli,
brandt-henglein}, but in every case it is confined to one domain and applies only to data already
presented in cyclic form; none reads it as the behaviour of an arbitrary system, and none folds a pure
term that merely generates the cycle.

% =====================================================================
\section{Conclusion}\label{sec:conclusion}
% =====================================================================
The behaviour of an equation morphism is read by tabling a one-step structure map, the structure map
itself the least fixpoint of the silent iteration. The unfolding diverges on a cyclic state; tabling
that map on first-order state identity folds a guarded cycle into a finite graph and decides an
unproductive cycle as $\bot$, both in finite time on the rational fragment. Head normalisation of
the pure $\lambda$-calculus is one instance, the one that shows a pure term can carry and compute over
cyclic data with nothing added to the calculus. The two parameters, the effect monad of the internal
process and the signature of the exposed layer, range over deterministic, non-deterministic,
probabilistic, and stateful processes alike.

% =====================================================================
\appendix
\section{The behaviour domain (Proposition~\ref{prop:cpo})}\label{app:domain}
\begin{proof}[Proof of Proposition~\ref{prop:cpo}]
\emph{Least element.} For any $t \in D$, $\mathrm{dom}(\bot_D) = \{\varepsilon\} \subseteq
\mathrm{dom}(t)$ and $\bot_D(\varepsilon) = \bot$, so $\bot_D \sqsubseteq t$ by
Definition~\ref{def:order}.

\emph{Directed completeness.} Let $\{t_i\}_{i\in I}$ be $\sqsubseteq$-directed. Define $t$ by
$\mathrm{dom}(t) = \bigcup_i \mathrm{dom}(t_i)$ and $t(p) = t_i(p)$ for any $i$ with
$t_i(p) \neq \bot$, else $t(p) = \bot$. This is well-defined: if $t_i(p) \neq \bot \neq t_j(p)$,
directedness gives $t_\ell \sqsupseteq t_i, t_j$, and a non-$\bot$ label is preserved upward
(Definition~\ref{def:order}), so $t_i(p) = t_\ell(p) = t_j(p)$. And $t \in D$: $\mathrm{dom}(t)$ is
prefix-closed as a union of prefix-closed sets, and arity-respecting because a non-$\bot$ label at $p$
agrees with some $t_i$, whose children of $p$ are exactly those the label prescribes, and $\bot$ has
no children. Finally $t = \bigsqcup_i t_i$: each $t_i \sqsubseteq t$, and any upper bound $u$ has
$t \sqsubseteq u$, since at $p$ with $t(p) \neq \bot$ the witnessing $t_i$ gives
$t(p) = t_i(p) = u(p)$. The $M$-structure at each node is completed $M$-pointwise, where $M A$ is a
pointed $\omega$-CPO by Definition~\ref{def:elgot}. In particular every $\omega$-chain has a least
upper bound.
\end{proof}

\section{The structure map (Proposition~\ref{prop:outwd})}\label{app:out}
\begin{proof}[Proof of Proposition~\ref{prop:outwd}]
Write $\Phi$ for the operator of Definition~\ref{def:out}, on $X \to M(F X)$ ordered pointwise over
the $\omega$-CPO structure of $M(F X)$ (Definition~\ref{def:elgot}).

\emph{Monotone and continuous.} $\Phi(g)(x) = \big([\,\eta\circ\mathrm{inl},\,g\,]^{*}\big)(\stepm(x))$
is built from $\stepm(x)$ (a constant in $g$) by Kleisli extension of $[\eta\circ\mathrm{inl}, g]$,
which depends on $g$ only through the $\mathrm{inr}$ branch and is $\omega$-continuous in $g$ because
$(-)^{*}$ and copairing are $\omega$-continuous (Definition~\ref{def:elgot}). Hence $\Phi$ is
$\omega$-continuous, and by Kleene $\outm = \bigsqcup_k \Phi^k(\lambda x.\,\bot)$ is its least
fixpoint.

\emph{Characterisation.} The iterates resolve layers by running silent steps: $\Phi^k(\bot)(x) \neq
\bot$ iff the silent process from $x$ exposes an $\mathrm{inl}$ layer within $k$ steps. \TODO{spell out
the step measure: an induction on $k$ matching ``$\Phi^k(\bot)(x) \neq \bot$'' with ``$x$ exposes a
layer within $k$ silent steps'', concluding $\outm(x) = \bot$ iff the silent process from $x$ exposes
no layer. The statement and both endpoints are fixed; only the bookkeeping of the measure remains.}
For $M = \Delaym$ this says $\outm(x) = \bot$ exactly when weak head reduction does not terminate
(Proposition~\ref{prop:shisout}).
\end{proof}

\section{Adequacy (Theorem~\ref{thm:adequacy})}\label{app:adequacy}
\begin{lemma}[Meaningless leaves]\label{lem:rootactive}
$\outm(x) = \bot$ iff the behaviour at $x$ is the $\bot$ leaf.
\end{lemma}
\noindent
By Proposition~\ref{prop:outwd}, $\outm(x) = \bot$ iff the silent process from $x$ exposes no layer.
The graph's $\bot$ leaf is by Definition~\ref{def:reach} exactly such a state, so the identification is
immediate.

\begin{proof}[Proof of Theorem~\ref{thm:adequacy}]
The structure map $\outm$ (Definition~\ref{def:out}) is a $\Beh$-coalgebra $X \to \Beh X$, $\Beh = MF$
(Definition~\ref{def:behfun}); the graph $(R(M),\outm)$ is its restriction to the reachable states. By
finality there is a unique $\Beh$-coalgebra morphism $\behav \colon X \to \nu\Beh$, determined layer
for layer by $\outm$: $\behav(x)$ exposes the layer $\outm(x)$ with children $\behav$ of the successor
states, or the $\bot$ leaf when $\outm(x) = \bot$ (Lemma~\ref{lem:rootactive}). Since
$\outm = \iter{\stepm}$ runs the silent steps of $\stepm$ (Proposition~\ref{prop:outwd}), $\behav$ is
the behaviour of $\stepm$ up to silent steps, and the graph $(R(M),\outm)$, whose unfolding is
$\behav(M)$, is its finite presentation.
\end{proof}

\section{Termination (Theorem~\ref{thm:termination})}\label{app:termination}
\begin{proof}[Proof of Theorem~\ref{thm:termination}]
Tabling descends only through reachable states (Definition~\ref{def:reach}): the children of an
exposed layer and, via the silent steps of $\outm$, the silent successors. State equality is decidable,
so the tabling key is testable.

\emph{Finite $\Rightarrow$ halts.} If $R(M)$ is finite, tabling $\outm$ visits a finite set and halts:
the silent collapse $\outm(x)$ saturates each finite silent orbit, and the reachable closure saturates
$R(M)$.

\emph{Halts $\Rightarrow$ finite.} Contrapositive: if $R(M)$ is infinite, the descent visits
infinitely many distinct states and the tabling never saturates.

\emph{Graph rational.} When $R(M)$ is finite, $(R(M),\outm)$ is a finite graph; its unfolding has
subtrees $\{\behav(y) : y \in R(M)\}$, finite, so it lies in $\rfp\Beh$
(Definition~\ref{def:rational}).

\emph{Two conditions.} The two halts above are independent (Remark~\ref{rem:tworationalities}): the
silent orbit of a state may be infinite while $R(M)$ is finite, or conversely.
\end{proof}

% =====================================================================
%% The body cites via \citeTODO (red draft markers), never \cite, so
%% \nocite{*} forces every references.bib entry into the printed list,
%% reproducing the former inline thebibliography under ACM formatting.
\nocite{*}
\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
